---
title: "CSP FINAL_PROJ"
author: "Vishal Balaji"
date: "2024-12-13"
output: html_document
---

# This study investigates the relationship between **marijuana use, gender, race**, and the levels of **depression** (categorized as "Several" or "Most") using the NHANES dataset. The primary goal is to understand how these variables influence depression risk and identify potential interaction effects.

```{r setup, include=FALSE}
library(NHANES)
df<-NHANES
library(dplyr)
library(ggplot2)
NHANES<-NHANES[!duplicated(NHANES$ID),]
library(table1)
library(lmtest)
library(car)
library(caret)
library(pROC)
library(corrplot)
library(knitr)
library(RVAideMemoire)
```

```{r, echo=TRUE}
# Set seed for reproducibility
set.seed(123) 
df_new <- NHANES[sample(nrow(NHANES), 6000), ]
df_new <-df_new[df_new$Depressed=='Several' | df_new$Depressed=='Most',]
# Select relevant variables for the study
variables <- c('ID', 'Age', 'Gender', 'Race1', 'Marijuana', 'Depressed')
df_new <- df_new[, variables]

df_new$depression_factor <- factor(df_new$Depressed, levels = c("Several", "Most"))

# Data Cleaning
# Handle missing values
df_new <- na.omit(df_new)
# Ensure appropriate data types
df_new$Gender <- factor(df_new$Gender, levels = c("male", "female"))
df_new$Race1 <- factor(df_new$Race1)
df_new$Marijuana <- as.numeric(df_new$Marijuana)

# Descriptive Statistics
# Table summarizing key variables
data_summary <- df_new %>%
  group_by(Gender, Race1) %>%
  summarise(
    Mean_Marijuana = mean(Marijuana, na.rm = TRUE),
    Count_Depressed = table(depression_factor)
  )
print(data_summary)

# Chi-squared test for Gender and Depression
chi_sq_test <- chisq.test(table(df_new$Gender, df_new$depression_factor))
print(chi_sq_test)

# Chi-squared test for Ethnicity and Depression
chi_sq_test_ethnicity <- chisq.test(table(df_new$Race1, df_new$depression_factor))
print(chi_sq_test_ethnicity)

# ANOVA for Marijuana Use across Depression Levels
anova_model <- aov(Marijuana ~ depression_factor, data = df_new)
summary(anova_model)

# Post-hoc test for ANOVA
post_hoc <- TukeyHSD(anova_model)
print(post_hoc)

# Residuals to find the direction of association
anova_residuals <- resid(anova_model)
ggplot(df_new, aes(x = depression_factor, y = anova_residuals)) +
  geom_boxplot(fill = "red", color = "black") +
  labs(title = "Residuals from ANOVA", x = "Depression Levels", y = "Residuals") +
  theme_minimal()

# Relative Risk/Odds Ratio Estimation
# Creating binary depression outcome (Several vs Most)
df_new$Depression_Binary <- ifelse(df_new$depression_factor == "Most", 1, 0)

# Logistic Regression for Odds Ratio
logistic_model <- glm(Depression_Binary ~ Marijuana + Gender + Race1, data = df_new, family = binomial)
summary(logistic_model)

# Calculate Odds Ratios
odds_ratios <- exp(coef(logistic_model))
print(odds_ratios)

# Visualization of Odds Ratios
library(ggplot2)
odds_ratios_df <- data.frame(
  Variable = names(odds_ratios),
  OddsRatio = odds_ratios
)

ggplot(odds_ratios_df[-1,], aes(x = Variable, y = OddsRatio)) +
  geom_bar(stat = "identity", fill = "grey") +
  labs(title = "Odds Ratios for Depression (Most vs Several)", x = "Variable", y = "Odds Ratio") +
  theme_minimal()

# Interaction Effects
interaction_model <- glm(Depression_Binary ~ Marijuana * Gender, data = df_new, family = binomial)
summary(interaction_model)

```



# <u><b>FINAL REPORT :</b></u>

# **Introduction**

This study investigates the relationship between **marijuana use, gender, race**, and the levels of **depression** (categorized as "Several" or "Most") using the NHANES dataset. The primary goal is to understand how these variables influence depression risk and identify potential interaction effects.

---

# **Data Gathering**

The dataset was sampled from **NHANES**, with 6,000 random observations. The study focused on participants who reported feeling depressed at the "Several" or "Most" level. The relevant variables selected for the analysis are:

- **ID**: Unique identifier for each participant  
- **Age**  
- **Gender**: Categorized as male or female  
- **Race1**: Categories include Black, Hispanic, Mexican, White, and Other  
- **Marijuana**: Numeric value indicating marijuana use  
- **Depressed**: Levels of depression ("Several" or "Most")  

A derived factor, **depression_factor**, was created to categorize depression levels.

---

# **Methods for Analysis**

### **Data Cleaning**

- Removed rows with missing values using `na.omit()`.  
- Converted **Gender** and **Race1** to factors, and **Marijuana** to a numeric type.

### **Descriptive Statistics**

- Grouped data by **Gender** and **Race1** to calculate the **mean marijuana use** and **count of depression levels**.

### **Chi-Square Tests**

- Tested the association between **Gender** and depression levels.  
- Tested the association between **Race1** and depression levels.

### **ANOVA**

- Analyzed differences in marijuana use across depression levels.  
- Conducted a **post-hoc Tukey HSD test** to determine specific group differences.

### **Logistic Regression**

- Modeled depression (binary outcome) based on **marijuana use, gender, and race**.  
- Calculated **odds ratios** to assess the risk factors.  
- Tested **interaction effects** between marijuana use and gender.

---

# **Analysis of Data**

## **Descriptive Statistics**

- Mean marijuana use was summarized for each combination of **Gender** and **Race1**.  
- The depression counts by group were also tabulated.

---

## **Chi-Square Tests**

### **Gender and Depression**

- **p-value = 0.3971** (Not significant)  
  No significant association between gender and depression levels.

### **Race and Depression**

- **p-value = 0.0962** (Not significant at 0.05 level)  
  No significant association between race and depression levels.

---

## **ANOVA for Marijuana Use**

- **p-value = 0.423** (Not significant)  
  No significant difference in marijuana use between participants reporting "Several" and "Most" depression levels.

### **Post-Hoc Test**

- No significant pairwise differences in marijuana use between the depression groups.

---

## **Logistic Regression**

### **Model Predictors**

- **Marijuana**: Estimate = -0.1502, p-value = 0.4289 (Not significant)  
- **Gender (female)**: Estimate = 0.1268, p-value = 0.4795 (Not significant)  
- **Race1 (Mexican)**: Estimate = -0.7180, p-value = 0.0305 (Significant)  

### **Odds Ratios**

- **Marijuana**: 0.86 (Not significant)  
- **Gender (female)**: 1.14 (Not significant)  
- **Race (Mexican)**: 0.49 (Significant) – Lower odds of reporting "Most" depression compared to the baseline group.

---

## **Interaction Effects**

- **Marijuana × Gender Interaction**:  
  **p-value = 0.1607** (Not significant)  
  The interaction between marijuana use and gender did not significantly affect depression levels.

---

# **Conclusion**

### **Key Findings**

1. **No significant associations** were found between **Gender** or **Race** and depression levels based on the Chi-Square tests.  
2. **Marijuana use** did not significantly differ across depression levels based on the ANOVA results.  
3. **Logistic regression** showed that **Mexican participants** had significantly lower odds of reporting "Most" depression compared to other racial groups.  
4. The interaction between **marijuana use and gender** was not statistically significant.

### **Summary**

The study did not find strong evidence linking marijuana use or gender to higher depression levels. However, race (specifically identifying as Mexican) showed a protective effect against severe depression.

---

# **Future Scope**

### **Larger Sample Size**

- Increase the dataset size to enhance statistical power and detect subtle effects.

### **Additional Variables**

- Include other mental health factors, socioeconomic status, education levels, and substance use patterns.

### **Longitudinal Analysis**

- Conduct a longitudinal study to observe changes in depression over time and how marijuana use influences these changes.

### **Mental Health Interventions**

- Explore how mental health services or interventions impact depression levels across different demographics.

### **Machine Learning Models**

- Use advanced models to capture non-linear relationships and interactions more effectively.

### **Psychosocial Factors**

- Investigate the impact of social support, stress, and life events on depression risk.

### **Interaction Studies**

- Examine interactions between marijuana use, gender, race, and other lifestyle factors (e.g., physical activity, diet).

This expanded scope can provide a more comprehensive understanding of the factors influencing depression.

---



















